---
title: "Ethiopia manuscript results"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: cerulean
    highlight: zenburn
    <!-- code_folding: hide -->
always_allow_html: true
---

##### 
<br>

```{r global_options, include=FALSE}

knitr::opts_chunk$set(fig.width = 12,
                      fig.height = 8,
                      fig.path = '../results/figs/test-summary/',
                      echo = TRUE, warning = FALSE, message = FALSE)

options(DT.options = list(pageLength = 100,
                          dom = 't',
                          filter = "top"))



out_figs_dir = "results/figs"
if( !exists(out_figs_dir) ) dir.create(out_figs_dir, recursive = TRUE)
```


## Setting up the workspace and loading data
```{r,  results = 'hide'}

# Packages
# ==============================================================================
library(tidyr)
library(ggplot2)
library(knitr)
library(DT)
library(stringr)
library(xtable)
library(hrbrthemes)
library(dplyr)
library(forcats)
library(finalfit)
library(dplyr)
library(stringr)
library(janitor)
library(lme4)
library(extrafont)
library(cowplot)
library(knitr)
library(parameters)
library(here)

# Load fonts
extrafont::loadfonts(quiet = TRUE)


# Load data
# ==============================================================================
bats_res = readRDS("data/Ethiopia_bat-info_test-summary.RDS")
bats_res$SiteName = trimws(gsub("-\\<Bat\\>|\\<Bat\\>", "", bats_res$SiteName))
bats_res$ConcurrentSamplingSite = bats_res$SiteName

tests = readRDS("data/Ethiopia_bat-test-info.RDS")
tests$SiteName = trimws(gsub("-\\<Bat\\>|\\<Bat\\>", "", tests$SiteName))
tests$Interface = ifelse(tests$ScientificName %in% "Rhinopoma hardwickii",
                      "Cave", "Building")

events = readRDS("data/Ethiopia_bat-sampling-events.RDS")

df = bats_res
df$Family = str_to_title(df$Family)
df$AgeClass = word(df$AgeClass, 1)

df$Interface = ifelse(df$ScientificName %in% "Rhinopoma hardwickii",
                      "Cave", "Building")

date_order = unique(format(as.Date(sort(df$EventDate), format = "%Y-%m-%d"),
                        "%Y-%B"))
df$EventMY = format(as.Date(df$EventDate, format = "%Y-%m-%d"), "%Y-%B")
df$EventMY = factor(df$EventMY, levels = date_order, ordered = TRUE)

tests = left_join(tests, df[ , c("AnimalID", "Family")], by = "AnimalID")

```

<br><br>

## Sampling events

```{r}

df %>%
    group_by(SiteName, EventDate, SeasonModelled) %>%
    tally(name = "No. of bats sampled") %>%
    group_by(EventDate) %>%
    arrange(SiteName, EventDate) %>%
    adorn_totals("row") %>%
    datatable(options = list(dom = 't'), rownames = FALSE)

```

<br><br>

## Viral families tested
```{r}

df %>%
    group_by(tests_done) %>%
    tally() %>%
    datatable(options = list(dom = 't'), rownames = FALSE)
    
```

<br><br>

## Species of bats tested

```{r}

df %>%
    select(Family, ScientificName) %>%
    group_by(Family, ScientificName) %>%
    tally() %>%
    adorn_totals("row") %>%
    datatable(options = list(dom = 't'), rownames = FALSE)

```

<br><br>

## Number of bats testing positive
Number of bats tested = `r nrow(df)`

Number of bats testing positive = `r sum(df$is_positive)`

Percent bats testing positive  = `r (sum(df$is_positive)/nrow(df)) * 100`

<br><br>

## Figure 1. Sampling sites

Made with ArcGIS 10.1

```{r, echo = TRUE}
out_file = file.path(here(), "manuscript_figures")
myimages = list.files(out_file, pattern = glob2rx("Figure-1*png"),
                      full.names = TRUE)
include_graphics(myimages)
```

<br><br>

## Table 1. Summary of  bats sampled and viral testing results

```{r}

df_ani =

    tests %>%
    select(SiteName, Interface, Family, ScientificName, AnimalID, VirusGroup) %>%
    group_by(SiteName, Interface, Family, ScientificName) %>%
    summarize(total_bats = length(unique(AnimalID))) %>%
    as.data.frame() %>%
    adorn_totals("row")



df_pos_ani =

    tests %>%
    select(SiteName, Family, ScientificName, AnimalID, VirusGroup) %>%
    filter( !is.na(VirusGroup)) %>%
    group_by(SiteName, Family, ScientificName) %>%
    summarize(pos_bats = length(unique(AnimalID))) %>%
    as.data.frame() %>%
    adorn_totals("row")


df_ani$pos_bats = paste0(
                      round(df_pos_ani$pos_bats/df_ani$total_bats * 100),
                      "% (", df_pos_ani$pos_bats, ")")
                      

df_spmn = 

    tests %>%
    select(SiteName, ScientificName, SpecimenType, SpecimenID) %>%
    mutate(SpecimenType = paste0("n_",
                                 gsub(" ", "_", SpecimenType))) %>%
    group_by(SiteName, ScientificName, SpecimenType) %>%
    summarize(total = length(unique(SpecimenID))) %>%
    as.data.frame() %>%
    spread(SpecimenType, total) %>%
    adorn_totals("row") %>%
    select(-SiteName)

df_pos_spmn =

    tests %>%
    select(SiteName, ScientificName, SpecimenType, SpecimenID, VirusGroup) %>%
    filter( !is.na(VirusGroup)) %>%
    mutate(SpecimenType = gsub(" ", "_", SpecimenType)) %>%
    group_by(SiteName, ScientificName, SpecimenType) %>%
    summarize(pos = length(unique(SpecimenID))) %>%
    as.data.frame() %>%
    spread(SpecimenType, pos) %>%
    mutate_all(~replace_na(., 0)) %>%
    adorn_totals("row") %>%
    select( -SiteName)

df_pos_spmn$oral_swab = paste0(
                            round(df_pos_spmn$oral_swab/df_spmn$n_oral_swab * 100),
                            "% (", df_pos_spmn$oral_swab, ")")

df_pos_spmn$rectal_swab = paste0(
                              round(df_pos_spmn$rectal_swab/df_spmn$n_rectal_swab * 100),
                              "% (", df_pos_spmn$rectal_swab, ")")


pos_viruses =

    df %>%
    select(ScientificName, pos_virus) %>%
    filter( !is.na(pos_virus)) %>%
    separate_rows(pos_virus, sep = "; ") %>%
    group_by(ScientificName, pos_virus) %>%
    tally() %>%
    mutate(viruses = paste0(pos_virus, " (", n, ")")) %>%
    group_by(ScientificName) %>%
    summarize(
        Viruses = paste0(sort(unique(viruses)), collapse = "<br/><br/>"))


bat_virus_info = left_join(df_ani, df_pos_spmn, by = "ScientificName")
bat_virus_info = left_join(bat_virus_info, pos_viruses, by = "ScientificName")

datatable(bat_virus_info, escape = FALSE, rownames = FALSE)

```

<br><br>

## Figure 2. Summary by sampling month, season, bat species, and sampling interface/location

<br>

#### Figure subcomponent 1: positives by sampling month and season


```{r}

p_df = df %>%
    group_by(EventMY, is_positive) %>%
    arrange(EventMY) %>%
    summarize(n = n()) %>%
    mutate(is_positive = ifelse(is_positive, "Virus Positive", "Virus Negative")) %>%
    mutate(is_positive = factor(is_positive, levels = c("Virus Positive", "Virus Negative")))

levels(p_df$EventMY) = gsub("-", "\n", levels(p_df$EventMY))

event_plot = ggplot(p_df, aes(x = EventMY, y = n, fill = is_positive, label = n)) +
    geom_bar(stat="identity", width = 0.5) +
    scale_fill_manual(values = c("#BD4F2C", "#2C42BD"), name = "Test result") +
    theme_ipsum(base_size = 12) +
    ylab("Number of bats") +
    xlab("") +
    theme(
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        legend.position = "top") +
    scale_y_continuous(expand = c(0,1))

p_season_df = df %>%
    group_by(EventMY, SeasonModelled) %>%
    arrange(EventMY) %>%
    summarize(n = n())

levels(p_season_df$EventMY) = gsub("-", "\n", levels(p_season_df$EventMY))

event_plot = event_plot + geom_text(data = p_season_df,
                       aes(label = SeasonModelled,
                           x = EventMY, y = n,
                           fill = NULL),
                       nudge_y = 2,
                       size = 3)


cowplot::save_plot("results/figs/event-plot.pdf", event_plot, base_height = 6)
cowplot::save_plot("results/figs/event-plot.png", event_plot, base_height = 6)

event_plot

```

<br><br>

#### Figure subcomponent 2: bat species tested
```{r}
bat_plot = df %>%
    select(ScientificName, is_positive) %>%
    group_by(ScientificName, is_positive) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    as.data.frame %>%
    mutate(ScientificName = factor(ScientificName,
                                   levels = c(
                                                "Neoromicia cf. somalica",
                                                "Mops midas",
                                                "Chaerephon pumilus",
                                                "Rhinopoma hardwickii"
                                            ))) %>%
    ggplot( aes(x = ScientificName, y= -n,
                fill = as.factor(-is_positive), width = 0.75)) +
    geom_bar(stat="identity") +
    scale_fill_manual(values = c("#BD4F2C", "#2C42BD"), name = "") +
    scale_x_discrete(name = "", position = "top") +
    scale_y_continuous(name = "No. of bats",
                       breaks = seq(0, -300, by = -50),  # y axis values (before coord_flip) 
                       labels = seq(0,  300, by =  50)) +
    coord_flip() +
    theme_ipsum(base_size = 12, plot_title_size = 12) +
    theme(
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position="none") +
    xlab("") +
    ylab("Number of bats") +
    labs(title = "Species tested")

cowplot::save_plot("results/figs/bat-plot.pdf", bat_plot, base_height = 3)

bat_plot

```

<br><br>

#### Figure subcomponent 3: Pie charts for species-event timeline

```{r}

sp_df = df %>%
    select(ScientificName, is_positive, EventDate) %>%
    group_by(ScientificName, is_positive, EventDate) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    mutate(sp_e = paste0(EventDate, "_", ScientificName)) %>%
    as.data.frame

sp_l = split(sp_df, sp_df$sp_e)

# ref:
# http://www.sthda.com/english/wiki/ggplot2-pie-chart-quick-start-guide-r-software-and-data-visualization

blank_theme = theme_minimal()+
    theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        panel.border = element_blank(),
        panel.grid=element_blank(),
        axis.ticks = element_blank(),
        plot.title=element_text(size=14, face="bold"),
        legend.position = "none"
    )


out_dir = "results/figs/event_pie"
if( !dir.exists(out_dir)) dir.create(out_dir)

invisible(lapply(sp_l, function(sp, out_dir) {
    
    if(nrow(sp) > 1) {
        sp_p = sp %>%
            ggplot(aes(x = "", y = n, fill = as.factor(-is_positive))) +
            geom_bar(width = 1, stat="identity") +
            scale_fill_manual(values = c("#BD4F2C", "#2C42BD"), name = "") +
            coord_polar("y", start=0) +
            theme_minimal() +
            blank_theme

        
    } else {
        sp_p = sp %>%
            ggplot(aes(x = "", y = n, fill = "#B9B7B6")) +
            geom_bar(width = 1, stat="identity") +
            scale_fill_manual(values = "#2C42BD", name = "") +
            coord_polar("y", start=0) +
            theme_minimal() +
            blank_theme

    }

    out_name_pdf = file.path(out_dir, paste0(gsub(" ", "_", unique(sp$sp_e)), ".pdf"))
        out_name_png = file.path(out_dir, paste0(gsub(" ", "_", unique(sp$sp_e)), ".png"))
        cowplot::save_plot(out_name_pdf, sp_p, base_height = .5)
        cowplot::save_plot(out_name_png, sp_p, base_height = .5)
        
}, out_dir))


```

```{r, echo = TRUE, out.width="15%"}
out_dir = file.path(here(), "results/figs/event_pie")
myimages = list.files(out_dir, pattern = ".png", full.names = TRUE)
include_graphics(myimages)
```

#### Final Figure 2 compiled with above subcomponents in Affinity Designer 1.8.3

```{r, echo = TRUE}
out_file = file.path(here(), "manuscript_figures")
myimages = list.files(out_file, pattern = glob2rx("Figure-2*png"),
                      full.names = TRUE)
include_graphics(myimages)
```

<br><br>


## Figure 3. Viral findings by bat species and sampling interface

<br><br>

#### Figure subcomponent 1: bar chart
```{r}

df_p =

    df %>%
    mutate(SeasonModelled = factor(SeasonModelled),
           SeasonModelled = fct_rev(SeasonModelled),
           ScientificName = ifelse(ScientificName %in% "Rhinopoma hardwickii",
                                   paste0(ScientificName, " (Cave)"),
                                   paste0(ScientificName, " (Building)")))
    

virus_plot =

    df_p %>%
    select(pos_virus, pos_virus_n, pos_virus_family,
           ScientificName, SeasonModelled) %>%
    drop_na() %>%
    mutate(pos_virus = gsub("; ", "\nand ", pos_virus)) %>%
    group_by(pos_virus, pos_virus_n, pos_virus_family,
             ScientificName, SeasonModelled) %>%
    summarize(n = n()) %>%
    arrange(pos_virus_family, desc(n)) %>%
    as.data.frame %>%
    mutate(pos_virus = factor(pos_virus, unique(pos_virus)),
           pos_virus = fct_rev(pos_virus)) %>%
    ggplot(aes(x = reorder(pos_virus, n), y=n, fill = ScientificName, label = n) ) +
    facet_wrap( ~ SeasonModelled ) +
    scale_fill_manual(values = c("#CC6677", "#44AA99", "#DDCC77", "#AA4499"),
                      name = "Bat species and\nsampling interface") +
    geom_bar(stat="identity") +
    ylim(0, 80) +
    # geom_text( size = 3, nudge_y = 4) +
    coord_flip() +
    theme_ipsum(base_size = 10) +
    theme(
        axis.text.y = element_text(size = 8),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position="right",
        legend.text = element_text(face = "italic")
    ) +
    xlab("") +
    ylab("Number of bats sampled") +
    labs(title = "Viruses detected in bats")


cowplot::save_plot("results/figs/virus-plot.pdf",
                   virus_plot, base_height = 6)

virus_plot

```

<br><br>

#### Figure subcomponent 2: Extra legend information

```{r}

df %>%
group_by(ScientificName, SeasonModelled, is_positive) %>%
    tally() %>%
    mutate(is_positive = ifelse(is_positive, "Positive", "Negative")) %>%
    pivot_wider(names_from = is_positive, values_from = n) %>%
    adorn_totals("col") %>%
    mutate_all(~replace_na(., 0)) %>%
    mutate(pos_percent = round((Positive/Total * 100), 2)) %>%
    mutate(info = paste0(pos_percent, "% (", Positive, "/", Total, ")")) %>%
    select(ScientificName, SeasonModelled, info) %>%
    pivot_wider(names_from = SeasonModelled, values_from = info) %>%
    datatable(options = list(dom = 't'))

```

<br><br>

#### Final Figure 3 compiled with above subcomponents in Affinity Designer 1.8.3
```{r, echo = TRUE}
out_file = file.path(here(), "manuscript_figures")
myimages = list.files(out_file, pattern = glob2rx("Figure-3*png"),
                      full.names = TRUE)
include_graphics(myimages)
```

<br><br>

## Viruses detected
```{r}

tbl = xtable(table(unlist(strsplit(df$pos_virus[ df$is_positive ], "; "))))
names(tbl) = "No. of bats testing positive for virus"
datatable(tbl)

```

<br><br>

## Coinfections
```{r}

df %>%
    filter( is_positive & pos_virus_n > 1 ) %>%
    select(SiteName, ScientificName, pos_virus) %>%
    group_by(SiteName, ScientificName, pos_virus) %>%
    summarize(n_bats = n()) %>%
    datatable(colnames = c("SiteName", "ScientificName", "Virus coinfection",
                           "No. of bats"))


```

<br><br>

## Virus detection by specimen type {.tabset .tabset-fade .tabset-pills}
### Breakdown of positive specimen types (at individual bat  level)

Rectal swabs were the most common specimen type to yield a positive virus sequence (92/99 bats), including 27 bats from whom sequences were confirmed in both rectal and oral swabs. In addition, seven bats had viruses detected only via their oral swabs.

```{r}

df %>%
    filter(is_positive) %>%
    group_by(pos_specimen_type) %>%
    tally() %>%
    adorn_totals("row") %>%
    datatable(options = list(dom = 't'))

```

<br><br>

### Positive specimen types for bats with > 1 virus detected
```{r}

ids = df$AnimalID[ which(df$pos_virus_n > 1) ]
multi_bats = unique(
                 tests[ tests$AnimalID %in% ids & !is.na(tests$VirusGroup),
                       c("AnimalID", "SpecimenType", "VirusGroup")]
             )


tbls = vector("list", 3)
for(i in seq_along(ids)) {
    tbls[[ i ]] = 
        multi_bats %>%
        filter(AnimalID %in% ids[ i ]) %>%
        datatable(options = list(dom = 't'))
}

tbls[[1]]
tbls[[2]]
tbls[[3]]
     

```

<br><br>

### Virus group by specimen type (at individual bat level)
```{r}

df %>%
    filter(is_positive) %>%
    select(pos_virus, pos_specimen_type) %>%
    group_by(pos_virus, pos_specimen_type) %>%
    tally() %>%
    as_data_frame() %>%
    spread(pos_specimen_type, n) %>%
    adorn_totals(c("row", "col")) %>%
    datatable(options = list(dom = 't'))

```

<br><br>

## Table 2. Viral prevalence in bats by sex, age class, season, species, and specimen type

_Fisher's exact tests_

```{r}

# Sex, age class, season, species, and interface
# ------------------------------------------------------------------------------
dependent = "is_positive"
explanatory = c("Sex",
                "AgeClass",
                "SeasonModelled",
                "ScientificName",
                "Interface")

t1 = df %>%
    summary_factorlist(dependent, explanatory, 
                       p = TRUE,
                       p_cat = "fisher") # Fisher's Exact test 
                       

# Specimen type
# ------------------------------------------------------------------------------
df_spm = df[ , c("AnimalID", "is_positive", "pos_specimen_type")]
df_spm = rbind(df_spm, df_spm)
df_spm$specimen = rep(c("oral swab", "rectal swab"), each = nrow(df))

df_spm$specimen_pos = mapply(grepl, df_spm$specimen, df_spm$pos_specimen_type)

t2 = df_spm %>%
    summary_factorlist("specimen_pos", "specimen", 
                       p = TRUE,
                       p_cat = "fisher") # Fisher's Exact test 


# Output
# ------------------------------------------------------------------------------
ft = rbind(t1, t2)
names(ft) = c("Group", "Variable", "FALSE", "TRUE", "p")

ft$pos = as.numeric(gsub("\\s*\\([^\\)]+\\)", "", ft$`TRUE`))
ft$neg = as.numeric(gsub("\\s*\\([^\\)]+\\)", "", ft$`FALSE`))
ft$pos_percent = round((ft$pos/(ft$pos + ft$neg) * 100), 1)

ft_out = ft %>%
    mutate(pos_percent = paste0(pos_percent, "% (", pos, "/", pos + neg, ")")) %>%
    select(Group, Variable, pos_percent, p)

knitr::kable(ft_out, align = c("l", "l", "r", "r"), caption = "Table 2")

```

<br><br>

## Exploring age/sex/season by species {.tabset .tabset-fade .tabset-pills}

### Rhinopoma
#### Crosstable
```{r}

# Crosstable 
dependent = "is_positive"
explanatory = c("Sex",
                "AgeClass",
                "SeasonModelled")

df %>%
    filter(ScientificName %in% "Rhinopoma hardwickii") %>%
    summary_factorlist(dependent, explanatory, 
                       p=TRUE,
                       add_dependent_label=TRUE,
                       p_cat = "fisher" # Fisher's Exact test 
                       ) %>%
    knitr::kable(align=c("l", "l", "r", "r", "r"))

```


<br><br>

### Chaerephon
#### Crosstable
```{r}

# Crosstable 
dependent = "is_positive"
explanatory = c(
                  "Sex",
                  "AgeClass",
                  "SeasonModelled")

df %>%
    filter(ScientificName %in% "Chaerephon pumilus") %>%
    summary_factorlist(dependent, explanatory, 
                       p=TRUE,
                       add_dependent_label=TRUE,
                       p_cat = "fisher" # Fisher's Exact test 
                       ) %>%
    knitr::kable(align=c("l", "l", "r", "r", "r"))

```


<br><br>

### Mops
#### Crosstable
```{r}

# Crosstable 
dependent = "is_positive"
explanatory = c(
                  "Sex",
                  # "AgeClass", # there are no subadult for Mops
                  "SeasonModelled")

df %>%
    filter(ScientificName %in% "Mops midas") %>%
    summary_factorlist(dependent, explanatory, 
                       p=TRUE,
                       add_dependent_label=TRUE,
                       p_cat = "fisher" # Fisher's Exact test 
                       ) %>%
    knitr::kable(align=c("l", "l", "r", "r", "r"))

```

<br><br>

### Neoromicia cf. somalica
#### Crosstable
```{r}

# Crosstable 
dependent = "is_positive"
explanatory = c(
                  "Sex"
                  # "AgeClass"
                  # "SeasonModelled"
              )

df %>%
    filter(ScientificName %in% "Neoromicia cf. somalica") %>%
    summary_factorlist(dependent, explanatory, 
                       p=TRUE,
                       add_dependent_label=TRUE,
                       p_cat = "fisher" # Fisher's Exact test 
                       ) %>%
    knitr::kable(align=c("l", "l", "r", "r", "r"))

```

<br>

---

<br>


## Multivariable logistic regression analyses
#### For _Rhinopoma hardwickii_, _Chaerephon pumilus_, and _Mops midas_

```{r, results = 'asis'}

df_filter = df %>%
    filter(! ScientificName %in% c("Neoromicia cf. somalica", "Mops midas"))

m1 = glm(is_positive ~ Sex + AgeClass + SeasonModelled + ScientificName,
           data = df_filter, family = binomial)

# summary(m1)
print_md(model_parameters(m1, exponentiate = TRUE))


df_filter = df %>%
    filter(! ScientificName %in% c("Neoromicia cf. somalica"))


m3 = glm(is_positive ~ Sex + SeasonModelled + ScientificName,
           data = df_filter, family = binomial)

# summary(m3)
print_md(model_parameters(m3, exponentiate = TRUE))

glm(is_positive ~ Sex + AgeClass + SeasonModelled + ScientificName,
    data = df_filter, family = binomial) %>%
    model_parameters(exponentiate = TRUE) %>%
    print_md()

dd =
    df %>%
    group_by(EventMY, EventDate, SeasonModelled, ConcurrentSamplingSite) %>%
    tally() %>%
    group_by(EventMY) %>%
    mutate(id = 1:n()) %>%
    arrange(EventDate)

df_filter = left_join(df_filter, dd, by = c("EventDate", "EventMY"))

```

